An introduction to wavelets using R

### 2022-01-09 text/14/wavelets.Rmd
More work on 2014 wavelets.Rmd

@@

![Figure 1. Graph of ecg signal](http://pmean.com/news/images/201312/wave01.png)

@@

@@

```{r graph-cosine-1, fig.height=1}
par(mar=c(2.6, 0, 0, 0))
t.unit <- (1:2048)/2048
cos1 <- +Re(ft.ecg[2])*cos(2*pi*t.unit)
plot(t.unit, cos1, type="l", axes=FALSE)
axis(side=1, pos=0, at=(0:4)/4, labels=(0:4)*512)
```

> graph.single("wave02.png",cos1,"cosine\nfunction",c(-0.35,0.35))
null device
1
> sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
> graph.single("wave03.png",sin1,"sine\nfunction",c(-0.35,0.35))
null device
1
> sum1 <- cos1+sin1
> graph.single("wave04.png",sum1,"Sum of\nsine and\ncosine",c(-0.35,0.35))
null device
1

Here's the first graph.

The graph of the cosine function shows a peak halfway through (1024) and a trough at the end (or equivalently at the beginning). This is actually flipped around from the normal cosine function, but recall that the multiplier front of the cosine function is negative.

The graph of the sine function shows a peak at 1/4 of the the way through (512) and a trough 3/4 of the way through (1536).

@@

> ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
> comp(round(2048*(atan(ft.ratio)+pi)/(2*pi)),which(sum1==max(sum1)),"Peak")
Peak
By hand 786
Using R 786
> comp(Mod(ft.ecg[2]),max(sum1),"Amplitude")
Amplitude
By hand 0.1931891
Using R 0.1931890

@@

> ft.comp2 <- rep(0,length(ft.ecg))
> ft.comp2[2] <- ft.ecg[2]
> comp2 <- Re(fft(ft.comp2,inverse=TRUE))
> mx.2 <- max(comp2)
> graph.single("wave05.png",comp2,"Second\nFourier\ncomponent",c(-0.35,0.35))
null device
1

>>>>

Here are some graphing functions I will need.

```{r graphing-functions}
graph_with_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
  par(mar=c(2.6, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
  text(1.1, x[2048], lab, adj=0.5)
}
graph_without_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
  par(mar=c(0, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  text(1.1, x[2048], lab, adj=0.5)
}
```

@@

```{r graph-ecg, fig.height=2}
graph_with_axis(ecg, "ecg\nsignal")
```

Figure 1. Graph of ecg signal

@@

```{r needle-plot, eval=FALSE}
par(mar=c(0,0,0,0))
plot(1:2048, Mod(ft.ecg), axes=FALSE)
plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
text(Re(ft.ecg), Im(ft.ecg), 1:2048)
```

@@

```{r graph-cosine-1, fig.height=1.33}
t.unit <- (1:2048)/2048
cos1 <- +Re(ft.ecg[2])*cos(2*pi*t.unit)
sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
add1 <- cos1+sin1

{
  graph_without_axis(cos1, "Cosine\ncomponent", limits=range(add1))
  graph_without_axis(sin1, "Sine\ncomponent", limits=range(add1))
  graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
}
```

The cosine function is a wave with a peak in the middle and a trough at the beginning (or equivalently, the end). This is actually the reverse of a cosine function, but only because the coefficient is negative.

The sine function is a wave with a peak a quarter of the way through and trough three quarters of the way through. A weighted sum of sine and cosine fuctions produces a trough at a different location. Here is about a third of the way through.

If you look at the original signal, the peak and trough roughly match the shape of the ecg curve, if you ignore the sudden spikes.

```{r compare-to-original-curve, fig.height=1.33}
{
  graph_without_axis(add1, "Combined\ncomponent")
  graph_without_axis(ecg, "Original\nsignal")
}
graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
peak <- (3*pi/2+Arg(ft.ecg[2]))/(2*pi)
amplitude <- Mod(ft.ecg[2])
segments(peak, 0, peak, amplitude)
segments(0, 0, 1, 0)
peak
atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))
atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))+pi
Arg(ft.ecg[2])
(which.max(add1)-1)/2047
amplitude
max(add1)
```

@@

```{r}
ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
round(2048*(atan(ft.ratio)+pi)/(2*pi))
which(add1==max(add1))
```

@@

```{r, fig.height=1.33}
ft.comp2 <- rep(0,length(ft.ecg))
ft.comp2[2] <- ft.ecg[2]
comp2 <- Re(fft(ft.comp2,inverse=TRUE))
{
  graph_without_axis(comp2, "Fourier\ncomponent", limits=range(add1))
  lines((1:2048)/2048, add1, col="red", lty="dotted")
}
```
